---
title: "NYPD Shooting Incident"
date: "`r Sys.Date()`"
output: html_document
---


## OVERVIEW

Two years of the pandemic worsened a mental-health crisis among the people in USA; over the same period, shootings have risen sharply. This has taken a deep toll on people, both those who have taken bullets and those in their orbit who have watched the aftermath.

This code ***imports, cleans, transforms and creates simple visualizations in R*** on the NYPD Shooting Incident Data to demonstrate different data science steps.

The data comes from the NYPD website that is available to the public to explore the nature of the activity/crime. The data is available for download [here](https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD).

## LIBRARIES  

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)  
library(lubridate)
library(ggplot2)
library(dplyr)
```
 

## IMPORTING THE DATA

The source is a csv file available on the NYPD website. This can be downloaded at the location specified above.

```{r Import}

NYPD_Shooting_Incidents_Data <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"  
nypd_csvData <- read_csv(NYPD_Shooting_Incidents_Data)  
```

## TIDYING AND TRANSFORMING

Tidying and transforming involves removing the "errors" as well as adapting it for our own use. The following letters are used to refer the code snippets involved.  

**a** - *OCCUR_DATE is standardized to date*   
**b** - *We will only be looking at data for the past 3 complete years*  
**c** - *Any columns not used for the purpose of this project is removed*  
**d** - *We will remove any rows that does not have a PERP_AGE_GROUP*  
 


```{r TidyTransform}  
# a
nypd_csvData <- nypd_csvData %>% mutate(OCCUR_DATE = mdy(OCCUR_DATE))

# b
nypd_csvData <- nypd_csvData %>% filter(OCCUR_DATE>="2019-01-01") 

# c  
nypd_csvData <- nypd_csvData %>% select(-OCCUR_TIME,-BORO,-PRECINCT,-JURISDICTION_CODE,-PERP_RACE,-PERP_SEX,-VIC_SEX,-VIC_RACE,-X_COORD_CD,-Y_COORD_CD,-Latitude,-Longitude,-Lon_Lat)

# d  
nypd_csvData <- nypd_csvData %>% drop_na(PERP_AGE_GROUP)  
#2219 rows were dropped of from a total of 4926 rows
```

## VISUALIZATION  

For the first plot, we will look at **perpetrator’s age within a category vs the number of incidents reported.**  

``` {r Perp_Age_Visual}  
nypd_csvData_by_perpAge <- nypd_csvData %>% group_by(PERP_AGE_GROUP)%>%
mutate(Incidents = n())  

nypd_csvData_by_perpAge_Plot <- nypd_csvData_by_perpAge %>% group_by(PERP_AGE_GROUP)%>%
summarize(Incidents = max(Incidents))  

nypd_csvData_by_perpAge_Plot  

nypd_csvData_by_perpAge_Plot %>% ggplot(aes(x=PERP_AGE_GROUP, y=Incidents))+ geom_bar(stat="identity", fill="steelblue")

```

For the second plot, we will look at **victim's age within a category vs the number of incidents reported.**  

```{r Victim_Age_Visual}
nypd_csvData_by_VicAge <- nypd_csvData %>% group_by(VIC_AGE_GROUP)%>%
mutate(Incidents = n())  

nypd_csvData_by_VicAge_Plot <- nypd_csvData_by_VicAge %>%
group_by(VIC_AGE_GROUP)%>%
summarize(Incidents = max(Incidents))  

nypd_csvData_by_VicAge_Plot  

nypd_csvData_by_VicAge_Plot %>% ggplot(aes(x=VIC_AGE_GROUP, y=Incidents))+ geom_segment(aes(x = VIC_AGE_GROUP, xend = VIC_AGE_GROUP, y = 0, yend = Incidents)) +
geom_point(color = "indianred4", size = 4)

```  

## ANALYZING DATA  

Let us now analyze the data further to examine the rate at which the incidents increased/decreased over the past year for the age group that had the highest number of victims (25-44; as identified using data visualization).  
Has the incidents been steady over the past year ?  
Which months had the highest incidents reported ?  
Is there a huge difference in number of incidents between the months ?  

```{r IncidentTrend2021}

nypd_csvData_Incidents <- nypd_csvData %>%
filter(year(OCCUR_DATE)==2021 & PERP_AGE_GROUP=="25-44")  

nypd_csvData_Incidents_plot <- nypd_csvData_Incidents %>%
mutate(OCCUR_Month = month(OCCUR_DATE))  

nypd_csvData_Incidents_plot <- nypd_csvData_Incidents_plot %>%
group_by(OCCUR_Month)%>%
summarise(Incidents = n(), Deaths = sum(STATISTICAL_MURDER_FLAG)) 

nypd_csvData_Incidents_plot <- nypd_csvData_Incidents_plot %>%
mutate(OCCUR_MonthName = month(OCCUR_Month,label = TRUE, abbr = TRUE))

nypd_csvData_Incidents_plot  

nypd_csvData_Incidents_plot %>% ggplot(aes(x=OCCUR_MonthName, y=Incidents, group = 1))+geom_line(color="darkslategrey")

```  

We have concluded that :  

The shooting incidents tend to increase between the months March and August and tends to decrease between September and February.  
May reported the highest shootings whereas January and February reported the least.  
There is almost 69% decrease in shootings in May vs January and February.  

## MODELING THE DATA  

A linear regression is a statistical model that analyzes the relationship between a response variable (often called y) and one or more variables and their interactions (often called x or explanatory variables). You make this kind of relationships in your head all the time, for example when you calculate the age of a child based on her height, you are assuming the older she is, the taller she will be.  

Similarly, we can assume that as the **number of incidents increases, the death rate also increases.**  

```{r LinearRegression}
mod <- lm(Deaths ~ Incidents, data = nypd_csvData_Incidents_plot)  

summary(mod)  

nypd_csvData_Incidents_Pred <- nypd_csvData_Incidents_plot %>%
mutate(pred = predict(mod))  

nypd_csvData_Incidents_Pred %>% ggplot() + geom_point(aes(x=Incidents, y=Deaths),color="black") + geom_point(aes(x=Incidents, y=pred),color = "red")  
```

## BIAS AREAS  

From the above model we can conclude that apart from a few outliers, our model somewhat predicts that as the incidents increases, the deaths also increases. Below are a few biases that could have affected our analysis :  

1. During tidy and transform, we had omitted rows that had NA values for the PERP_AGE_GROUP. This might have affected the model as more data means more accurate results.    

2. The victim's age group considered for analysis, 25-44 is a vast age group. It may be possible that the victims at the lower end of age group may be at a better position to defend/survive the shooting than someone at the higher end. Underlying health conditions can contribute towards survival rate, which is usually observed for higher age groups.  







 
















